Type theories as quotient inductive-recursive types

### 2025-02-20 tex/main.tex
Fix typos and minor revisions

\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}./}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko for useful discussion.}%optional

@@

In this paper, we show that the development can be streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s).
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and precursors by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: pattern matching definitions with overlapping patterns.
%
This is helpful to avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees~\cite{Cockx2017}.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with QIITs alone).
%

@@

\subparagraph*{Contributions}
This paper explores the use of QIIRTs and definitions by overlapping patterns to define type theories in intensional type theory. Our contributions are:
\begin{itemize}
  \item We give definitions of the parallel substitution calculus and type theories with a universe and $\Pi$-types as a QIIRT (\cref{sec:QIIRTs}).
  \item We show how elimination principles for reasoning about the type theory can be derived, and use it to construct its standard model (\cref{sec:models}).
  \item We compare our definition to definitions using quotient inductive-inductive types, and prove that the two definitions are actually equivalent (\cref{thm:equivalence}).
\end{itemize}

@@

\subsection{Inductive-recursive types}
An inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined using an elimination rule which says that $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of~$A$.
%For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\suc\,n)$, where the latter might refer to the already computed value $f(n)$.
%
The type is first defined inductively by listing its constructors, and then recursive functions can be defined out of it.

In contrast, an inductive-recursive type (IRT), introduced by Dybjer~\cite{Dybjer2000} and studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}, is an inductive type $A$ defined \emph{simultaneously} with a recursively defined function $F : A \to D$ to some fixed (possibly large) type $D$.
The prototypical example is Tarski universe: a type $U$ of codes for small types is defined with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents. Crucially, the code
\[
  \pi : (a : U) \to (T(a) \to U) \to U
  \quad\text{with}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T(b\, x)
\]
for dependent function types makes sense only because we have access to the function $T$.
%

An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P(x)$ by giving values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$, but the recursive function $F : A \to D$ is \emph{not} defined using this elimination rule, but part of the definition of the type itself.
%
In particular, the elimination rule can be used to define endofunctions on $A$, but the recursive function  $F : A \to D$ must target a pre-existing type $D$.
%
In this work, we will consider an extension of \emph{generalised} IRTs, where $F : A \to A$ is also allowed.
%
Such IRTs are accepted in \Agda and were previously used~\cite{Danielsson2006} to represent type theory in type theory.
%
%Their general metatheory would be interesting future work.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT in \cref{fig:QIT-join-list}.
\begin{figure}[b]
\begin{minipage}{.35\textwidth}
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
\end{alignat*}
\end{minipage}
\begin{minipage}{.64\textwidth}
\begin{alignat*}{3}
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}
\end{minipage}
\caption{The quotient inductive type of join lists}
\label{fig:QIT-join-list}
\end{figure}
%\vskip\belowdisplayskip\noindent
A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list. We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$ from \cref{fig:QIT-join-list}, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'
\]
which states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.
%In this work, we will consider quotient inductive-inductive-recursive types, generalising QIITs and QITs.

@@

A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any overlapping clauses must eventually compute to the same term, i.e.\ the definition must be confluent.
Cockx shows that if a definition of $f$ is
\begin{enumerate}
  \item \emph{structurally recursive}~\cite{Abel2002}, i.e.\ arguments of recursive calls are structurally smaller, and
  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, the terms $\sigma t_1$ and $\sigma t_2$ have the same normal form, 
\end{enumerate}
then $f$ is confluent --- essentially a variant of Newman's Lemma~\cite{Newman1942}.
We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ used in the analysis of local confluence a \emph{local peak}.

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts such as the one below.
\begin{example}\label{ex:pi2-comp}
  Given any substitution $\sigma : \Sub\;\Gamma\; \Delta$ and a non-empty substitution $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, the first term of the composite $(\sigma; \tau)$ should intuitively be equal to the first term of $\tau$ after applying the substitution $\sigma$, i.e.\ the identity
  \[
    \pi_2\,(\sigma ; \tau) = \sub{(\pi_2\,\tau)}{\sigma}
  \]
  apparently holds and can checked by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, \sub{(\pi_2\,\tau)}{\sigma})
    = \sub{(\pi_2\,\tau)}{\sigma}.
  \end{equation}
  Yet, the left-hand side is a term of type $\sub{A}{\pi_1\,(\sigma;\tau)}$, but the other is $\sub{\sub{A}{\pi_1\,\tau}}{\sigma}$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} \sub{(\pi_2\,\tau)}{\sigma}$ for some $p: \sub{A}{\pi_1(\sigma; \tau)} = \sub{\sub{A}{\pi_1\tau}}{\sigma}$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    (\sub{A}{\pi_1(\sigma; \tau)}, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} (\sub{\sub{A}{\pi_1\,\tau}}{\sigma}, \sub{(\pi_2\,\tau)}{\sigma})
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

@@

One possibility is to define the substitution calculus as a QIIRT, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer1999}, since the codomain of the recursive function is the inductively defined typed itself.
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define $[\blank]_{\Ty} \blank : \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma$ recursively as follows:
\begin{alignat}{3}
%\subTy{\blank}{\blank} &: \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \nonumber \\
\subTy{\UU}{\sigma}         & \defid \UU                                   \label{eq:type-sub-at2-1} \\
\subTy{A}{\idS}             & \defid A                                     \label{eq:type-sub-at2-2} \\
\subTy{A}{\sigma ; \tau}    & \defid \subTy{\subTy{A}{\tau}}{\sigma} \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be elaborated to 
\[
\subTy{\UU}{\emptysub}         \judgeq \UU, \qquad
\subTy{\UU}{\sigma, t}         \judgeq \UU, \qquad\text{and}\qquad 
\subTy{\UU}{\pi_1\,\sigma}     \judgeq \UU,
\]
Thus in this case we have again lost the desired strict identity $\subTy{\UU}{\sigma} \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

With this definition of type substitution, other equality constructors can be introduced without any $\transp$, and we can give the updated constructors the types below.
\begin{alignat*}{5}
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) {} && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), \sub{t}{\sigma} \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)          {} && =^{\Tm\,\Gamma\, A} &&  t \\
  [\idS]t         & : \implicit{\Gamma, A, t} && {\sub{t}{\idS}}          && =^{\Tm\,\Gamma\,A}     {} && t \\
  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}  && {\sub{t}{\sigma ; \tau}} && =^{\Tm\,\Gamma\,\sub{A}{\sigma ; \tau}} {} && {\sub{\sub{t}{\tau}}{\sigma}}
\end{alignat*}

@@

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
%
In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by the reduction
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
\end{alignat*}
Yet, this naive change breaks local confluence!
The problem is that there is now additional interaction between types and terms.
% Observe that
The local peak
\(
  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
\)
cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
%

>>>>

\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko for useful discussion.}%optional

@@

In this paper, we show that the development can be streamlined by combining quotient inductive-inductive types and inductive-recursive types, called \emph{quotient inductive-inductive-recursive types} (QIIRTs).
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and precursors by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: pattern matching definitions with overlapping patterns.
%
This is helpful to avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees~\cite{Cockx2017}.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with QIITs alone).
%

@@

\subparagraph*{Contributions}
This paper explores the use of QIIRTs and definitions by overlapping patterns to define type theories in intensional type theory. Our contributions are:
\begin{itemize}
  \item We give definitions of the parallel substitution calculus and type theories with a universe and $\Pi$-types as a QIIRT (\cref{sec:QIIRTs}).
  \item We show how elimination principles for reasoning about the type theory can be derived, and use it to construct its standard model (\cref{sec:models}).
  \item We compare our definition to definitions using quotient inductive-inductive types, and prove that the two definitions are actually equivalent (\cref{thm:equivalence}) in type theory.
\end{itemize}

@@

\subsection{Inductive-recursive types}
An inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined using an elimination rule which says that $f$ is defined by its values $f(\cc_i\,x)$ on all of the constructors $\cc_i$ of~$A$.
%For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\suc\,n)$, where the latter might refer to the already computed value $f(n)$.
%
The type is first defined inductively by listing its constructors, and then recursive functions can be defined out of it.

In contrast, an inductive-recursive type (IRT), introduced by Dybjer~\cite{Dybjer2000} and studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}, is an inductive type $A$ defined \emph{simultaneously} with a recursively defined function $F : A \to D$ to some fixed (possibly large) type $D$.
The prototypical example is Tarski universe: a type $U$ of codes for small types is defined with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents. Crucially, the code
\[
  \pi : (a : U) \to (T\,a \to U) \to U
  \quad\text{with}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T\,(b\, x)
\]
for dependent function types makes sense only because we have access to the function $T$.
%

An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P\,x$ by giving values $f(\cc_i\,\overrightarrow{x})$ on all of the constructors $\cc_i$ of $A$, but the recursive function $F : A \to D$ is \emph{not} defined using this elimination rule, but part of the definition of the type itself.
%
In particular, the elimination rule can be used to define endofunctions on $A$, but the recursive function  $F : A \to D$ must target a pre-existing type $D$.
%
In this work, we will consider an extension of \emph{generalised} IRTs, where $F : A \to A$ is also allowed.
%
Such IRTs are accepted in \Agda and were previously used~\cite{Danielsson2006} to represent type theory in type theory.
%
%Their general metatheory would be interesting future work.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT below.
%\begin{figure}[b]
\begin{minipage}{.4\textwidth}
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
\end{alignat*}
\end{minipage}
\begin{minipage}{.58\textwidth}
\begin{alignat*}{3}
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}
\end{minipage}
%\caption{The quotient inductive type of join lists}
%\label{fig:QIT-join-list}
%\end{figure}
\vskip\belowdisplayskip\noindent
A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list and replaces equality constructors $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'.
\]
This constructor states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.
%In this work, we will consider quotient inductive-inductive-recursive types, generalising QIITs and QITs.

@@

A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any overlapping clauses must eventually compute to the same term, i.e.\ the definition must be confluent.
Cockx shows that if a definition of $f$ is
\begin{enumerate}
  \item \emph{structurally recursive}~\cite{Abel2002}, i.e.\ arguments of recursive calls are structurally smaller, and
  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, the terms $\sigma t_1$ and $\sigma t_2$ have the same normal form, 
\end{enumerate}
then $f$ is confluent --- essentially a variant of Newman's Lemma~\cite{Newman1942}.
We call a diagram $\sigma t_2 \Leftarrow f\,(\sigma\overrightarrow{p_1}) \Rightarrow \sigma t_1$ used in the analysis of local confluence a \emph{local peak}.

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts such as the one below.
\begin{example}\label{ex:pi2-comp}
  Given any substitution $\sigma : \Sub\;\Gamma\; \Delta$ and a non-empty substitution $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, the first term of the composite $(\sigma; \tau)$ should intuitively be equal to the first term of $\tau$ after applying the substitution $\sigma$, i.e.\ the identity
  \[
    \pi_2\,(\sigma ; \tau) = \sub{(\pi_2\,\tau)}{\sigma}
  \]
  apparently holds and can be checked by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, \sub{(\pi_2\,\tau)}{\sigma})
    = \sub{(\pi_2\,\tau)}{\sigma}.
  \end{equation}
  Yet, the above identity does not even make sense, since their types do not match --- the left-hand side is a term of type $\sub{A}{\pi_1\,(\sigma;\tau)}$, but the other is $\sub{\sub{A}{\pi_1\,\tau}}{\sigma}$.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} \sub{(\pi_2\,\tau)}{\sigma}$ for some $p: \sub{A}{\pi_1(\sigma; \tau)} = \sub{\sub{A}{\pi_1\tau}}{\sigma}$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    (\sub{A}{\pi_1(\sigma; \tau)}, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} (\sub{\sub{A}{\pi_1\,\tau}}{\sigma}, \sub{(\pi_2\,\tau)}{\sigma})
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right-hand side.

@@

One possibility is to define the substitution calculus as a QIIRT, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer1999}, since the codomain of the recursive function is the inductively defined typed itself.
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define $\subTy{\blank}{\blank} : \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma$ recursively as follows:
\begin{alignat}{3}
%\subTy{\blank}{\blank} &: \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \nonumber \\
\subTy{\UU}{\sigma}         & \defid \UU                                   \label{eq:type-sub-at2-1} \\
\subTy{A}{\idS}             & \defid A                                     \label{eq:type-sub-at2-2} \\
\subTy{A}{\sigma ; \tau}    & \defid \subTy{\subTy{A}{\tau}}{\sigma} \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be elaborated to 
\[
\subTy{\UU}{\emptysub}         \judgeq \UU, \qquad
\subTy{\UU}{\sigma, t}         \judgeq \UU, \qquad\text{and}\qquad 
\subTy{\UU}{\pi_1\,\sigma}     \judgeq \UU,
\]
Thus in this case we have again lost the desired strict identity $\subTy{\UU}{\sigma} \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

With this definition of type substitution, other equality constructors can be introduced without any $\transp$, and we can give the updated constructors the types below.
\begin{alignat*}{5}
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) {} && =^{\Sub\,\Gamma\,(\Theta, A)} {} &&  (\sigma ; \tau), \sub{t}{\sigma} \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)          {} && =^{\Tm\,\Gamma\, (\subTy{A}{\sigma})} {} &&  t \\
  [\idS]         & : \implicit{\Gamma, A, t} && {\sub{t}{\idS}}          && =^{\Tm\,\Gamma\,A}      {} && t \\
  [;]            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}  && {\sub{t}{\sigma ; \tau}} && =^{\Tm\,\Gamma\,(\subTy{A}{\sigma ; \tau})} {} && {\sub{\sub{t}{\tau}}{\sigma}}
\end{alignat*}

@@

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Ty\;\Gamma} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
%
In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by the reduction
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
\end{alignat*}
Yet, this naive change breaks local confluence!
The problem is that there is now additional interaction between types and terms.
% Observe that
The local peak
\(
  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
\)
cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
%
